---
title: "Tables of Most Common Rare Metabolic Diseases"
author: "Gary Koplik"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

```{r, message = F, warning = F}

# clear old code to confirm replicability
rm(list = ls() )

# load libraries
library(ggplot2)
library(dplyr)
library(knitr)

```


### Methodology

I started with the subset of patients that at one point were assigned an ICD-10 code of a rare disease (from the mappings).

I then filtered these data to only keep the rows containing a rare disease ICD-10 code (remember, we have restructured the data such that there is only one ICD-10 code per row).

I finally further filtered these data such that the combination of patient ID and ICD-10 code was unique / a primary key in this final data frame (for example, if a patient has cystic fibrosis with the specified ICD-10 code in a visit in October of 2012 and shows the same code in October of 2013, then one of these rows will be deleted). I did this to avoid double counting patients with respect to determining prevalence of rare diseases in this data set. *Note, we can go back later to look **only** at patients who are assigned a rare disease code more than once, which is likely a more robust indication of whether a patient has a rare disease vs possibly having a rare disease*.

```{r, message = F, warning = F}

#### data set up ####

# load data sets

# full size data sets
load("../data/rarePatients.Rdata")
rare <- rareData
rm(rareData)
```

### Frequency Table with All ICD-10 Codes for Rare Diseases


This table shows the frequency of all ICD-10 codes in our rare disease file (from the National Library of Medicine). I also include the ICD-10 code description as well as the name of the disease it maps to in our mapping file.

```{r, fig.width = 10, warning = F, message = F, cache = T}

# read in table
icd10 <- read.table("../data/ORDR_ICD10_FULL.txt", sep = "\t", quote = "",
                    stringsAsFactors = F, header = T)

icd10 <- icd10 %>%
  # only keep primary terms for disease name
  filter(ORDR_TTY == "PT") %>%
  dplyr::select("code" = ICD10CM_CODE,
                "disease_num" = ï..ORDR_ID,
                "disease_name" = ORDR_NAME,
                "code_description" = ICD10NAME)

# rename for compatibility with old code
disease_codes <- icd10
# disease_codes <- base::rbind(disease_codes_icd9, disease_codes_icd10)

# keep only the rows with a rare icd code
rare_people <- dplyr::filter(rare, rare$Code %in% disease_codes$code)

# keep unique icd codes within each PAT_ID
rare_people$unique <- !duplicated(data.frame(rare_people$Code,
                                             rare_people$PAT_ID))

rare_representation_unique <- dplyr::filter(rare_people,
                                            unique == T)

# from stackoverflow question 27422229

# avoid duplicates of icd10 codes
icd10_unique <- unique(icd10[c("code", "code_description")])
icd10_unique <- icd10_unique %>% dplyr::rename("Code" = code)

icd10 <- icd10 %>% dplyr::rename("Code" = code)
rare_representation_unique_icd10 <- rare_representation_unique

# get sorted frequencies
tab <- data.frame(rev(sort(table(rare_representation_unique_icd10$Code))))
tab <- tab %>%
  dplyr::rename("Code" = Var1) %>%
  left_join(icd10, by = "Code") %>%
  dplyr::select("ICD-10 Code" = Code,
                "Frequency" = Freq,
                "Rare Disease Name" = disease_name,
                "ICD Code Description" = code_description) %>%
  # keep only metabolic diseases (E70 - E88)
  filter(substr(`ICD-10 Code`, 1, 1) == "E") %>%
  filter(as.numeric(substr(`ICD-10 Code`, 2, 3)) %in% c(70:88))
```

Number of ICD-10 Codes: **`r nrow(tab)`**

Total Number of Patients: **`r sum(tab[2])`**

```{r}

kable(tab)

```



### Frequency Table of ICD-10 Codes that Each Only Map to Exactly 1 Rare Disease

Here, I specifically looked at only ICD-10 codes that uniquely mapped to exactly one "rare" disease in our mapping files.


```{r, fig.width = 10, message = F, warning = F, cache = T}

# using the codes that map to 1 disease in each (9 and 10) data set only

# create a list of rare diseases to use to subset out just rare diseases
# specifically icd codes that map to only 1 rare disease

# see ./code_uniqueness.R for more details


load("../data/icd10_1to1.Rdata")
icd10 <- icd10_1to1
rm(icd10_1to1)

# rename for compatibility with old code
disease_codes <- icd10

# keep only the rows with a rare icd code
rare_people <- dplyr::filter(rare, rare$Code %in% disease_codes$code)

# keep unique icd codes within each PAT_ID
rare_people$unique <- !duplicated(data.frame(rare_people$Code,
                                             rare_people$PAT_ID))

rare_representation_unique <- dplyr::filter(rare_people,
                                            unique == T)

# rename code to Code for later stuff lining up
icd10 <- icd10 %>% dplyr::rename("Code" = code)
rare_representation_unique_icd10 <- rare_representation_unique %>%
  # add in details of icd codes
  left_join(icd10)


tab <- data.frame(rev(sort(table(rare_representation_unique_icd10$Code))))
tab <- tab %>%
  dplyr::rename("Code" = Var1) %>%
  left_join(icd10, by = "Code") %>%
  dplyr::select("ICD-10 Code" = Code,
                "Frequency" = Freq,
                "Rare Disease Name" = disease_name,
                "ICD Code Description" = code_description) %>%
  # keep only metabolic diseases (E70 - E88)
  filter(substr(`ICD-10 Code`, 1, 1) == "E") %>%
  filter(as.numeric(substr(`ICD-10 Code`, 2, 3)) %in% c(70:88))


```

Number of ICD-10 Codes: **`r nrow(tab)`**

Total Number of Patients: **`r sum(tab[2])`**


```{r}
kable(tab)

```

$\hspace{4ex} _\blacksquare$




